---
title: "Mushroom Classifier"
output: html_notebook
---

```{r Setup, message = F, include = F}
library(tidyverse)


mushrooms <- readr::read_csv("mushrooms.csv", 
col_types = as.list(rep("c", 23)))
# Some parsing failures, need to revisit. 

```


# Introduction

This is a dataset of mushrooms and 23 characteristics, and 23 species of mushroom ($N=8124$). The dataset is available on this repo, but is also available with more information from [here](https://www.kaggle.com/uciml/mushroom-classification/data).

**Dataset feature summary:**

- `class` : classes: edible=e, poisonous=p

- `cap-shape` :  bell=b,conical=c,convex=x,flat=f, knobbed=k,sunken=s

- `cap-surface` : fibrous=f,grooves=g,scaly=y,smooth=s 

- `cap-color`: brown=n,buff=b,cinnamon=c,gray=g,green=r,pink=p,purple=u,red=e,white=w,yellow=y

- `bruises`: bruises=t,no=f

- `odor`:  almond=a,anise=l,creosote=c,fishy=y,foul=f,musty=m,none=n,pungent=p,spicy=s

- `gill-attatchment` : attached=a,descending=d,free=f,notched=n

- `gill-spacing` : close=c,crowded=w,distant=d

- `gill-size` : broad=b,narrow=n

- `gill-color` : black=k,brown=n,buff=b,chocolate=h,gray=g, green=r,orange=o,pink=p,purple=u,red=e,white=w,yellow=y

- `stalk-shape` : enlarging=e,tapering=t

- `stalk-root` : bulbous=b,club=c,cup=u,equal=e,rhizomorphs=z,rooted=r,missing=?

- `stalk-surface-above-ring` : fibrous=f,scaly=y,silky=k,smooth=s

- `stalk-surface-below-ring` : fibrous=f,scaly=y,silky=k,smooth=s

- `stalk-color-above-ring` : brown=n,buff=b,cinnamon=c,gray=g,orange=o,pink=p,red=e,white=w,yellow=y

- `stalk-color-below-ring` : brown=n,buff=b,cinnamon=c,gray=g,orange=o,pink=p,red=e,white=w,yellow=y

- `veil-type` : partial=p,universal=u

- `veil-color` : brown=n,orange=o,white=w,yellow=y

- `ring-number` : none=n,one=o,two=t

- `ring-type` : cobwebby=c,evanescent=e,flaring=f,large=l,none=n,pendant=p,sheathing=s,zone=z

- `spore-print-color` : black=k,brown=n,buff=b,chocolate=h,green=r,orange=o,purple=u,white=w,yellow=y

- `population` : abundant=a,clustered=c,numerous=n,scattered=s,several=v,solitary=y

- `habitat` : grasses=g,leaves=l,meadows=m,paths=p,urban=u,waste=w,woods=d

# Exploratory Data Analysis



```{r }
# Might do something with this later.
# mushrooms %>% 
#   pivot_longer(cols = colnames(mushrooms)[2:23]) %>% 
#   group_by(class, name, value) %>% 
#   count() %>% 
#   ungroup() %>% 
#   group_by(class) %>% 
#   ggplot(aes(name,value, fill = n)) +
#   geom_tile() +
#   coord_flip()+
#   facet_wrap(~class)

# Classes Look balanced
mushrooms %>% 
  group_by(class) %>% 
  count()
```


# Edible or Not?


